---
output:
  html_document:
    fig_caption: yes
    keep_md: yes
---
An adaptive MCMC via covariance adjustment
====

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8)

```


```{r, echo = F}
set.seed(123)
```


## Create test data
As a first step, we create some test data that will be used to fit our model. Let’s assume a linear relationship between the predictor and the response variable, so we take a linear model and add some noise.

```{r test_data}
# Load libraries
rm(list=ls(all=TRUE))
library(devtools)
source_url('https://raw.githubusercontent.com/florianhartig/Cookbook/master/Plotting/Correlation/CorrelationDensityPlotWithIpanel.r')
library(coda)
library(MASS)

# Define variables
trueA <- 5
trueB <- 0
trueSd <- 10
sampleSize <- 31

# Creating uncentered predictor variable, will lead to correlation
# between intercept and slope in the posterior
x <- (-(sampleSize-1)/2):((sampleSize-1)/2) + 20
y <-  trueA * x + trueB + rnorm(n=sampleSize,mean=0,sd=trueSd)

 
# Plot data
plot(x,y, main="Test Data")
```


## Defining the statistical model 
The following definitions are standard for a linear regression, if you don't understand why, read ( ?)



```{r}
# Likelihood function
likelihood <- function(param){
  a = param[1]
  b = param[2]
  sd = param[3]
  
  pred = a*x + b
  singlelikelihoods = dnorm(y, mean = pred, sd = sd, log = T)
  sumll = sum(singlelikelihoods)
  return(sumll)  
}

# Prior distribution
prior <- function(param){
  a = param[1]
  b = param[2]
  sd = param[3]
  aprior = dunif(a, min=0, max=10, log = T)
  bprior = dnorm(b, sd = 5, log = T)
  sdprior = dunif(sd, min=0, max=30, log = T)
  return(aprior+bprior+sdprior)
}
```
<br />

## The MCMC

Now comes the mcmc to sample from the posterior. We are using a Metropolis-Hastings MCMC. I you are uncertain how this works, first read http://theoreticalecology.wordpress.com/.

In this previous version, the proposal function was simlply creating independent normal draws $proposalfunction <- function(param){return(rnorm(3,mean = param, sd= c(0.1,0.5,0.3))) }$ .

We change now to a multivariate normal version that will be adapted later.

```{r}

sig = diag(x = c(0.1,0.5,0.3), nrow=3, ncol=3)

proposalfunction <- function(param){
  return(mvrnorm(1,mu = param, Sigma= sig))
  }

```
Here is the mcmc. I adapted the code slightly to allow for running the MCMC for a while, stopping, and continuing with the MCMC, which will be useful for the adaptation


```{r Metropolis_algorithm}
 ################ Metropolis algorithm ################
run_metropolis_MCMC <- function(iterations){
  startindex = nrow(chain)
  chain = rbind(chain, array(dim = c(iterations,3)))
  for (i in startindex:(startindex+iterations-1)){
    proposal = proposalfunction(chain[i,])
    
    probab = exp(likelihood(proposal)+ prior(proposal) - likelihood(chain[i,])- prior(chain[i,]))
    if (runif(1) < probab){
      chain[i+1,] = proposal
    }else{
      chain[i+1,] = chain[i,]
    }
  }
  return(chain)
}
```

But first of all, we will just run the MCMC like before and check the convergence as explained in http://theoreticalecology.wordpress.com/2011/12/09/mcmc-chain-analysis-and-convergence-diagnostics-with-coda-in-r/


```{r convergence}

# Checking the convergence
chain = array( c(4,2,8), dim = c(1,3))

# Running a non-adapted analysis with deliberately bad proposal function 

# Settings for the proposal covariance matrix
sig = diag(x = c(1,1,1), nrow=3, ncol=3)

chain1 = run_metropolis_MCMC(10000)
chain2 = run_metropolis_MCMC(10000)
combinedchains = mcmc.list(mcmc(chain1), mcmc(chain2))
plot(combinedchains)
gelman.diag(combinedchains)
gelman.plot(combinedchains)

```


<br />

#### Adaptation steps 
We use the samples obtained already to adjust the proposal for exaplanations why the scaling factor of 2.38^2/d is optional (see references)



```{r Adaptation_steps }

chain = array( c(4,2,8), dim = c(1,3))
chain = run_metropolis_MCMC(2000)
sig = 2.38^2 / 3 * cov(chain) 
chain = run_metropolis_MCMC(2000)
sig = 2.38^2 / 3 * cov(chain) 
chain = run_metropolis_MCMC(2000)
sig = 2.38^2 / 3 * cov(chain) 
chain = run_metropolis_MCMC(2000)
sig = 2.38^2 / 3 * cov(chain) 
chain = run_metropolis_MCMC(2000)
sig = 2.38^2 / 3 * cov(chain) 

 
chain = array( c(4,2,8), dim = c(1,3))
chain1 = run_metropolis_MCMC(10000)
chain2 = run_metropolis_MCMC(10000)
combinedchains = mcmc.list(mcmc(chain1), mcmc(chain2))
plot(combinedchains)
gelman.diag(combinedchains)
gelman.plot(combinedchains)

```

<br />
<br />

## References 

Rosenthal, J. S. (2011) Optimal proposal distributions and adaptive MCMC. Handbook of Markov Chain Monte Carlo, CRC Press.
