---
title: "Model Averaging"
author: "FlorianHartig"
date: "15 May 2015"
output: html_document
---

Based on an example provided by Jörn Pagel. 

```{r, echo=F, warning=F, message=F}
set.seed(123)
rm(list=ls(all=TRUE))
library(rjags)
library(effects)
```


## Creation of test data

```{r, fig.width=7, fig.height=7}
a <- 5
b <- 10
c <- 0.1
sigma <- 10

x <- -15:15
y <- a * x + b + x^2 * c + rnorm(31,0,sd = sigma)
plot(x,y)
```



## Bayesian analysis of this model (in Jags)

```{r}
  # 1) Model definition exactly how we created our data 
  model1 = textConnection("
    model{
      # Likelihood
      for(i in 1:i.max){
      y[i] ~ dnorm(mu[i],tau)
      mu[i] <- a*x[i] + b
      }
      # Prior distributions
      a ~ dnorm(0,0.001)
      b ~ dnorm(0,0.001)
      tau <- 1/(sigma*sigma)
      sigma ~ dunif(0,100)
    }
  ")

  model2 = textConnection("
    model{
      # Likelihood
      for(i in 1:i.max){
      y[i] ~ dnorm(mu[i],tau)
      mu[i] <- a1*x[i] + a2*x[i]*x[i] + b
      }
      # Prior distributions
      a ~ dnorm(0,0.001)
      b ~ dnorm(0,0.001)
      tau <- 1/(sigma*sigma)
      sigma ~ dunif(0,100)
    }
  ")
  
  # 2) Set up a list that contains all the necessary data (here, including parameters of the prior distribution)
  Data = list(y = y, x = x, i.max = length(y))

  # 3) Specify a function to generate inital values for the parameters
  inits.fn <- function() list(a1 = rnorm(1), a2 = rnorm(1), b = rnorm(1), sigma = runif(1,1,100))

```

Running the model

```{r, fig.width=7, fig.height=7}
  # Compile the model and run the MCMC for an adaptation (burn-in) phase
  jagsModel <- jags.model(file= model1, data=Data, init = inits.fn, n.chains = 3, n.adapt= 1000)

  # Specify parameters for which posterior samples are saved
  para.names <- c('pD', 'deviance', "a1","a1","b","sigma")

  # Continue the MCMC runs with sampling
  Samples <- coda.samples(jagsModel, variable.names = para.names, n.iter = 5000)
  
  # Plot the mcmc chain and the posterior sample for p
  plot(Samples)
```








http://pluto.coe.fsu.edu/mcmc-hierMM/

mcmixWAIC <- function (Y,pi,mu,sigma) {
  ppd <- mcmixPost(Y,pi,mu,sigma)  # R x J matrix
  lppd <- sum(log(apply(ppd,2,mean)))
  pWAIC1 <- 2*sum(log(apply(ppd,2,mean))-apply(log(ppd),2,mean))
  pWAIC2 <- sum(apply(log(ppd),2,var))
  WAIC1 <- -2*(lppd-pWAIC1)
  WAIC2 <- -2*(lppd-pWAIC2)
  c(lppd=lppd, pWAIC1=pWAIC1, WAIC1=WAIC1, pWAIC2=pWAIC2, WAIC2=WAIC2)
}

mcmixDIC <- function (Y,pi,mu,sigma) {
  ppd <- mcmixPost(Y,pi,mu,sigma)  # R x J matrix
  sppd <- apply(log(ppd),1,sum) ## Sum across observations.
  pi.bayes <- matrix(apply(pi,2,mean),1)
  mu.bayes <- matrix(apply(mu,2,mean),1)
  sigma.bayes <- matrix(sqrt(apply(sigma^2,2,mean)),1) # Take average of
                                        # variance, not sd
  lppd.bayes <- sum(log(mcmixPost(Y,pi.bayes,mu.bayes,sigma.bayes)))
  pDIC <- 2*(lppd.bayes - mean(sppd))
  pDICalt <- 2*var(sppd)
  DIC <- -2*(lppd.bayes-pDIC)
  DICalt <- -2*(lppd.bayes-pDICalt)
  c(lppd=mean(sppd), lppd.bayes=lppd.bayes, pDIC=pDIC, DIC=DIC,
    pDICalt=pDICalt, DICalt=DICalt)
}



## Further readings and applications 


http://conradstack.blogspot.de/2012/10/reversible-jump-markov-chain-monte.html
http://stats.stackexchange.com/questions/4328/reversible-jump-mcmc-code-matlab-or-r

https://www.ceremade.dauphine.fr/~xian/BCS/

http://winbugs-development.mrc-bsu.cam.ac.uk/rjmcmc.html


### RJ-MCMC implementations 

http://www.inside-r.org/packages/cran/auteur/docs/rjmcmc.bm
http://www.inside-r.org/packages/cran/geiger/docs/rjmcmc.bm






---
**Copyright, reuse and updates**: By Florian Hartig. Updates will be posted at https://github.com/florianhartig/LearningBayes. Reuse permitted under Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License
