Statistical concepts, at the example of the coin flip model
===


Assume we had 10 trials, 7 success, unknown coin, want to do inference to learn about the coin

```{r}
rm(list=ls(all=TRUE))
trials = 10
success = 7
```

# GETTING THE MLE ESTIMATE

The idea of maximum likelihood estimation (MLE) is to look for the set of parameters that would, under the given model assumption, lead to the highest probability to obtain the observed data. In our case we have only one parameter, the probability of success per flip. Let's plot this for different values and look for the maximum.  

```{r}
# parameters to check
parametervalues <- seq(0,1,0.001)
# get the probability for all parametervalues
likelihood <- dbinom(success,trials,parametervalues)

# plot results
plot(parametervalues, likelihood, type = "l")
legend("topleft", legend = c("Likelihood", "maximum"), col = c("black", "red"), lwd = 1)
MLEEstimate <- parametervalues[which.max(likelihood)]
abline(v=MLEEstimate, col = "red")
```


## Confidence intervals

MLE confidence intervals are constructed by comparing likelihood values of single parameters against each other. We have the MLE estimate as null hypothesis and ask ourselves how likely it is that the true parameter is a different one. By cutting at 0.05 probability, we construct the frequentist CI interval. The test statistics that can be used to do this are discussed, e.g., in https://onlinecourses.science.psu.edu/stat504/node/39. The result for a 1-parameter model is that the CI is at a log likelihood difference of 1.96

```{r}
plot(parametervalues, likelihood, type = "l")
legend("topleft", legend = c("Likelihood", "maximum", "CI"), col = c("black", "red", "green"), lwd = 1)
MLEEstimate <- parametervalues[which.max(likelihood)]
abline(v=MLEEstimate, col = "red")

confidence.level <- log(max(likelihood)) -1.96
leftCI <- parametervalues[which.min(abs(log(likelihood[1:which.max(likelihood)]) - confidence.level))]
abline(v=leftCI, col = "green")
rightCI <- parametervalues[which.min(abs(log(likelihood[which.max(likelihood):length(likelihood)]) - confidence.level)) + which.max(likelihood) -1]
abline(v=rightCI, col = "green")


```


# GETTING THE P-VALUE FOR FAIR COIN

want to get p-value for a smaller or equal result (1-tailed) given a fair coin p(k<=kobs|H0:p=0.5). Basically, we want the sum over the red bars

```{r}
barplot(dbinom(0:10, 10, 0.5), col = c(rep("grey", success ), rep("red", 11-success)))
line(pbinom(0:10,trials,prob = 0.5, lower.tail = F))
```


We can get this with the cummulative distribution function in R

```{r}
pValue <- pbinom(success-1,trials,prob = 0.5, lower.tail = F)
pValue
pValue < 0.05
```

but it is a bit tricky, because depeding on which side one wants to test, you have to add a -1 (because of the definition of the cummulative). So, it's instructive to do it that way, but easier in practice to use the binom.test, which calculates the same values 

```{r}
binom.test(7,trials,0.5) # two sided 
binom.test(7,trials,0.5, alternative="greater") # testing for greater
binom.test(7,trials,0.5, alternative="less") # testing for less
```

## Multiple testing

Imagine we do 20 times a test, but we have no effect.

```{r}
data= rbinom(20,10,0.5)
pValue <- pbinom(data,trials,prob = 0.5, lower.tail = F)
pValue
max(pValue < 0.05) == 1

pValueAdjusted <- p.adjust(pValue, method = "hochberg")
pValueAdjusted
max(pValueAdjusted < 0.05) == 1
```

## Long-term frequencies of errors 

imagine we have a fair coin, and do 1000 experiments with 10 draws, looking at the p-value, note that we get to the right 5% assumptotically, but not for small data. The fact that it is not exactly 5% is an artifact that appears for discrete distributions (<8 gets acceted, >= 8 rejected, the perfect cut would be intbetween but as we have only discrete options that's not possible to choose)


```{r}
par(mfrow=c(2,1))
trials = 10
outcome <- rbinom(1000, trials, 0.5)
hist(pbinom(outcome,trials,0.5), breaks = 100, main = "10 trials, 11 p-values" )
sum(pbinom(outcome,trials,0.5) < 0.05)


trials = 10000
outcome <- rbinom(1000, trials, 0.5)
hist(pbinom(outcome,trials,0.5), breaks = 100, main = "10000 trials, p-values flat"  )
sum(pbinom(outcome,trials,0.5) < 0.05)
```

remember: in general, if you choose an alpha level of 5%, and you have absolutely random data, you should get 5% false positives (type I error) assymptotically, and the distribution of p-values in repeated experiments will be flat. 

Also note: we are free to choose the null-hypothesis as we want. What would you do if you null hypothesis is that a coint should have an 0.8 proability of head?

# The BAYESIAN ESTIMATE

Remember for Bayes p(M|D) = p(D|M) * p(M) / P(D), and we can show that p(D) is just the integral over p(D|M) * p(M)

We had already calculated p(D|M), so we just need to define p(M), the prior

for flat prior p(M) = flat = 1
See http://en.wikipedia.org/wiki/Jeffreys_prior , section on Bernoulli trial, to understand that this is not neccesarily the best uninforative choice, but it is simple at any rate

```{r}
prior <- rep(1,1001)
posterior <- likelihood * prior / sum(likelihood * prior) * length(parametervalues)

plot(parametervalues, posterior, col = "darkgreen", type = "l")
lines(parametervalues, likelihood)
lines(parametervalues, prior, col = "red" )
legend("topright", c("likelihood", "prior", "posterior"), col = c("black", "red", "green"), lwd = 1 )
```

you see that likelihood and posterior have the same shape. However, this is only because I chose a flat prior. There is still a difference, however, namely that the posterior is normalized, i.e. will integrate to one. It has to be, because we want to interpret it as a pdf, while the likelihood is not a pdf. Let's look at the same example for an informative prior


```{r}
prior <- dnorm(parametervalues, mean = 0.5, sd = 0.1)
posterior <- likelihood * prior / sum(likelihood * prior) * length(parametervalues)

plot(parametervalues, posterior, col = "darkgreen", type = "l")
lines(parametervalues, likelihood)
lines(parametervalues, prior, col = "red" )
legend("topright", c("likelihood", "prior", "posterior"), col = c("black", "red", "green"), lwd = 1 )
```

you can see that the likelihood moves the posterior away from the prior, but not by much. try the same think with more data, but the same ratio, i.e. change to 30 trials, 9 success


# Bayesian CI

calculated via central 95% probability

```{r}
plot(parametervalues, posterior, col = "darkgreen", type = "l")
lines(parametervalues, likelihood)
lines(parametervalues, prior, col = "red" )
legend("topright", c("likelihood", "prior", "posterior"), col = c("black", "red", "green"), lwd = 1 )

cumPost <- cumsum(posterior) / (length(posterior) -1)

leftCI <- parametervalues[which.min(abs(cumPost - 0.025))]
abline(v=leftCI, col = "darkgreen", lty = 2, lwd = 2)
leftCI <- parametervalues[which.min(abs(cumPost - 0.975))]
abline(v=leftCI, col = "darkgreen", lty = 2, lwd = 2)
```







