---
title: "Priors in Bayesian analysis"
author: "FlorianHartig"
date: "9 May 2015"
output:
  html_document:
    keep_md: yes
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=5, fig.height=5)
```


```{r, echo=F}
set.seed(123)
```

Priors are key to 


# Priors on scale parameters

## Informative choices


## Uninformative choices

http://stats.stackexchange.com/questions/61928/choosing-non-informative-priors?rq=1


## Bayesian Regularization via mildly informative priors 

Kyung, M.; Gill, J.; Ghosh, M.; Casella, G. et al. (2010) Penalized regression, standard errors, and Bayesian lassos. Bayesian Analysis, 5, 369-411.

http://stats.stackexchange.com/questions/95395/ridge-regression-bayesian-interpretation?rq=1


# Priors on variance parameters

## Regressions


## Hierarchical models

http://stats.stackexchange.com/questions/6493/weakly-informative-prior-distributions-for-scale-parameters







---
**Copyright, reuse and updates**: By Florian Hartig. Updates will be posted at https://github.com/florianhartig/LearningBayes. Reuse permitted under Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License

