---
title: "Linear regression with Jags"
author: "Florian Hartig"
date: "30 Jul 2014"
output:
  html_document:
    keep_md: yes
---


```{r, echo=F, warning=F, message=F}
set.seed(123)
rm(list=ls(all=TRUE))
library(rjags)
library(runjags)
library(effects)
library(R2jags)
```


## Creation of test data

```{r, fig.width=5, fig.height=5}
a <- 5
b <- 10
sigma <- 10

x <- -15:15
y <- a * x + b + rnorm(31,0,sd = sigma)
plot(x,y)
```

## Non-Bayesian analysis of this model

```{r, fig.width=5, fig.height=5}
fit <- lm(y ~ x)
summary(fit)
plot(allEffects(fit, partial.residuals = T))
```



## Bayesian analysis of this model (in Jags)

```{r}
  # 1) Model definition exactly how we created our data 
  modelCode = "
    model{
      # Likelihood
      for(i in 1:i.max){
      y[i] ~ dnorm(mu[i],tau)
      mu[i] <- a*x[i] + b
      }
      # Prior distributions
      a ~ dnorm(0,0.001)
      b ~ dnorm(0,0.001)
      tau <- 1/(sigma*sigma)
      sigma ~ dunif(0,100)
    }
  "
  
  # 2) Set up a list that contains all the necessary data (here, including parameters of the prior distribution)
  Data = list(y = y, x = x, i.max = length(y))

  # 3) Specify a function to generate inital values for the parameters
  inits.fn <- function() list(a = rnorm(1), b = rnorm(1), sigma = runif(1,1,100))

```


Running the model with rjags

```{r, fig.width=7, fig.height=7}
  # Compile the model and run the MCMC for an adaptation (burn-in) phase
  jagsModel <- jags.model(file= textConnection(modelCode), data=Data, init = inits.fn, n.chains = 3, n.adapt= 1000)

  # Specify parameters for which posterior samples are saved
  para.names <- c("a","b","sigma")

  # Continue the MCMC runs with sampling
  Samples <- coda.samples(jagsModel, variable.names = para.names, n.iter = 5000)
  
  # Plot the mcmc chain and the posterior sample for p
  plot(Samples)
```

convergence check

```{r}
gelman.diag(Samples)
```

```{r}
summary(Samples)
```

predictions (not very elegant)

```{r, fig.width=5, fig.height=5}
plot(x,y)
sampleMatrix <- as.matrix(Samples)
selection <- sample(dim(sampleMatrix)[1], 1000)
for (i in selection) abline(sampleMatrix[i,1], sampleMatrix[i,1], col = "#11111105")

```

# Running the model with runjags

```{r}
runJagsResults <- run.jags(model=modelCode, monitor=c("a","b","sigma"), data=Data, n.chains=2, method="rjags", inits=inits.fn)
plot(runJagsResults)
```


# Running the model with R2jags

```{r,  fig.width=14, fig.height=14}
R2JagsResults <- jags(data=Data, inits=inits.fn, parameters.to.save=c("a","b","sigma"), n.chains=2, n.iter=5000, model.file=textConnection(modelCode))

plot(R2JagsResults)
print(R2JagsResults)

```

Change to coda standard format

```{r, fig.width=14, fig.height=14}
R2JagsCoda <- as.mcmc(R2JagsResults)
plot(R2JagsCoda)
summary(R2JagsCoda)
```


---
**Copyright, reuse and updates**: By Florian Hartig. Updates will be posted at https://github.com/florianhartig/LearningBayes. Reuse permitted under Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License
