---
title: "Benchmark functions for MCMC, optimization and Sensitivity Analysis"
author: "FlorianHartig"
date: "28 Apr 2015"
output:
  html_document:
    keep_md: yes
---

This is an overview of functions that are useful for benchmarking MCMC, optimization and SA.


# The soobench package

The soobench (Single Objective Optimization Benchmark Functions) package provides a Collection of different single objective test functions useful for benchmarks and algorithm development.

```{r, message=F, warning=F}
library(soobench)
```


See http://cran.r-project.org/web/packages/soobench/index.html

# Tests in the FME package

The FME package has a vignette with tests of the FME mcmc implementations. See

```{r, message=F, warning=F}
library(FME)
#vignette("FMEmcmc")
```

An example of this vignette (from Laine ,2008) is a banana-shaped function is created by distorting a two-dimensional Gaussian distribution,
with mean = 0 and a covariance matrix τ with unity variances and covariance of 0.9:

```{r}
Banana <- function (x1, x2) {
  return(x2 - (x1^2+1))
}

# We need a function that estimates the probability of a multinormally distributed vector

pmultinorm <- function(vec, mean, Cov) {
  diff <- vec - mean
  ex <- -0.5*t(diff) %*% solve(Cov) %*% diff
  rdet <- sqrt(det(Cov))
  power <- -length(diff)*0.5
  return((2.*pi)^power / rdet * exp(ex))
}

# The target function returns -2 *log (probability) of the value

BananaSS <- function (p){
  P <- c(p[1], Banana(p[1], p[2]))
  Cov <- matrix(nr = 2, data = c(1, 0.9, 0.9, 1))
  -2*sum(log(pmultinorm(P, mean = 0, Cov = Cov)))
}
```

# Other Benchmarks

* mlbench: Machine Learning Benchmark Problems  http://cran.r-project.org/web/packages/mlbench/


